### 2018-03-09 submission/manuscript.Rmd
discussion edits with pat comments in mind

Although we expected that the full OTU models would perform the best at classifying individuals with and without carcinomas, our observations suggest that both the full and significant OR taxa models performed equally well. These results suggest that lower level classification to species and strain may not add extra useful information with respect to prediction models. This has been suggested in previous literature where metagenomics did not perform better than 16S rRNA gene sequencing data at classifying individuals with normal colons and those with carcinomas [@Hannigan2017]. One possible reason as to why lower level classification may not result in better models is that the communities are patchy and higher level taxonomic information pools some of this patchiness, allowing for better prediction models. There may also be a fair bit of data redundancy within models that utilize more of the community. An example of this redundancy would be when we trained models on one study and tested it on the other studies and the AUCs of the models created with the select OR taxa performed as well as full taxa models [Figure 6B].

Our observations also suggest that a small collection of taxa can classify disease as well as full OTU-based models but that these taxa individually perform quite poorly [Figure 3]. This result supports the contention that there might be redundancy of function amongst the taxa included in the significant OR models. As an example, multiple different microbes could be similarily stimulating the activation of inflammatory pathways and by doing so exacerbate disease progression. Multiple reports within the literature have found that different bacteria, such as *Escherichia coli* and *Fusobacterium nucleatum*, can similarily worsen inflammation in mouse models of tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @kostic_fusobacterium_2013]. Although the inflammatory taxa were patchy in their importance and presence across studies those that were not typically associated with inflammation were consistently important for every study [Figure 6]. The loss of these taxa (*Ruminococcus* and *Clostridium XI* in stool and *Dorea* and *Blautia* in unmatched tissue) is particularly interesting because many are commonly thought to be beneficial due to their involvement in production of short chain fatty acids [@Venkataraman2016; @Herrmann2018; @Reichardt2017].

The adenoma models as a whole performed poorly in classifying individuals with and without adenomas. This outcome is not inconsistent with what has been published previously [@hale_shifts_2017; @Sze2017]. However, the modeling results are at odds with results obtained in Baxter, *et al.* [@baxter_microbiota-based_2016]. There are some major differences between the models generated in this meta-analysis and what was used in Baxter, *et al*. First, the prevoius report's models investigated the classification of lesions (individuals with adenoma or carcinoma) and not adenoma alone. The Baxter, *et al.* models also contained Fecal Immunoglobulin Test data while our meta-analysis models only contained 16S rRNA gene sequencing data. Although being able to classify individuals with adenomas is important, the classification of advanced adenomas is a more clinically meaningful diagnostic tool (i.e. those that are at highest risk of progressing to a carcinoma). It is possible that we might have been able to detect differences in the bacterial community if advanced adenomas were separated from adenomas but that data was not available for the majority of studies analyzed. Additionally, the initial changes to the bacterial community could be focal to where the initial adenoma develops and would not be easily assessed with a fecal sample.

Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- tissue associated microbial communities. The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors. Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S3]. *Fusobacterium* was not consistently identified across studies and this could be due to both a small number of studies and a small sample size within these studies. The majority of the colon tissue-based results were also study specific with many of the top 10 taxa being present only in a single study. Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying (e.g. *Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc. [@Salter_contamination_2014]). The low bacterial biomass of tissue samples coupled with potential contamination and small sample sizes could explain why these results seem to be more sporadic than the stool results.

One important caveat to this study is that even though genera associated with certain species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp. *gallolyticus* were not identified, it does not necessarily mean that these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017]. There are reports that *Bacteroides fragilis*, positive for the enterotoxigenic gene, are found at specific locations along the colon but the samples we were able to use in this meta-analysis could not identify these types of differences [@Purcell2017]. Additionally, since we are limited in our aggregation of the data to the genus level, it is not possible to clearly delineate which species are contributing to overall disease progression. Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp. *gallolyticus*. As an example, the stool-based full community models consistently identified the *Bacteroides* taxa to be important model components across studies. This suggests that even though *Bacteroides* may not increase the OR of individuals having an adenoma or carcinoma and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC.

Despite these limitations to the findings, meta-analyses are a useful tool in microbiota research because they can both validate existing research and make new discoveries by pooling many independent investigations together. Yet, it is still difficult to perform these studies because of inaccessible 16S sequencing data, missing or vague metadata (e.g. which samples are carcioma and which are not), varying sequence quality, and multiple small data sets. Better attention to these specific problems could help to increase the reproducibility and replicability of microbiota studies and make it easier to perform these crucial meta-analyses. Moving forward, meta-analyses will be important tools to help aggregate and find commonalities across studies when investigating the microbiota in the context of a specific disease and more are needed [@Sze2016; @Walters2014; @Finucane2014; @shah_leveraging_2017].

By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in carcinoma tumors. Although further validation of the biomarkers presented here need to be undertaken, the replicability of the AUC of a specific collection of taxa across multiple studies suggests a strong potential for the use of the microbiota as a risk stratification tool for individuals with carcinomas.

@@

***Random Forest classification analysis:*** 
To classify individuals as having normal colons or tumors, we built Random Forest classification models for each dataset and comparison using taxa with significant ORs (after multiple comparison correction), all taxa, or OTUs. Because no taxa were identified as having a significant OR associated with adenomas using stool samples or tissue samples, classification models based on OR data were not constructed to classify individuals as having normal colons or adenomas. Within the training dataset, 10-fold cross validation (5-fold cross validation for small datasets) was used to build a model that was then evaluated on the testing set. For the models constructed using the taxa with significant ORs, the default mtry setting was used to train the model and this model was tested on the other datasets in the meta-analysis. The reported AUC values are the AUCs for the application of the model on the test sets. For the OTU-based models, the dataset was split into training (80% of samples) and testing (20%) sets and 10-fold cross validation (5-fold cross validation for small datasets) on the training set was used to generate the model for the testing set. The original 80/20 split and fitting was repeated 100 times and the average AUC from these 100 repeats was reported. The Mean Decrease in Accuracy (MDA), a measure of the importance of each taxon to the overall model was used to rank the taxa used in each model. For all models, the default setting used was $\sqrt{p}$, where p is all the variables used in the respective model.

>>>>

We performed a meta-analysis to identify and validate microbiome-base biomarkers that could be used to classify individuals as having normal colons or colonic tumors using fecal or tissue samples. To our surprise, Random Forest classification models constructed to differentiate individuals with normal colons from those with carcinomas using a subset of the community performed well relative to models constructed using the full communities. When we applied the models trained on each dataset to the other datasets in our study, we found that the models trained using the subset of the communities performed better than those using the full communities. These models were trained using data in which sequences were assigned to bacterial taxa using a classifier that typically assigned sequences to the family or genus level. When we attempted to improve the specificity of the classification by using an OTU-based approach the resulting models performed as well as those constructed using coarse taxonomic assignments. These results are significant because they strengthen the growing literature indicating a role of the microbiome in tumorigenesis [@zackular_gut_2013] and as a potential tool as a non-invasive diagnostic and for assessing risk of disease and recurrence [@baxter_microbiota-based_2016; @Sze2017].

These results suggested that fine scale classification of sequences into OTUs does not improve our classification models. This has been suggested in previous literature where shotgun metagenomic data did not perform better than 16S rRNA gene sequencing data in classifying individuals with normal colons and those with carcinomas [@Hannigan2017]. We hypothesize that fine scale classification may not result in better classification because distribution of microbiota between individuals is patchy. In contrast, models using coarser taxonomic assignments will pool the fine scale diversity and resulting in less patchiness and better classification. Furthermore, the ability of models trained using a subset of the community to outperform those using the full community when testing the models on the other datasets may also be a product of the patchiness of human-associated microbiota. The models based on the 8 taxa that had significant ORs used taxa that were found in every study and tended to have higher relative abundances. Similar to the OTU-based models, those models based on the full community taxonomy assignments were still sensitive to the patchy distribution of taxa. Regardless, it is encouraging that a collection of  `r length(crc_OR_stool$measure)` taxa could reliably classify individuals as having carcinomas considering the differences in cohorts, DNA extraction procedures, regions of the 16S rRNA gene, and sequencing methods.

When used separately to classify individuals with carcinomas, the taxa with significant ORs could not reliably classify individuals [Figure 3]. This result further supports the hypothesis that carcinoma-associated microbiota have a patchy distribution. Two individuals may have had the same classification, based on the relative abundance of different populations within this group of `r length(crc_OR_stool$measure)` taxa. Although these results only reflect associations with disease, it is tempting to hypothesize that the patchiness represents distinct mechanisms of exacerbating tumorigenesis or that multiple taxa have the same mechanism of exacerbating tumorigenesis. For example, strains of *Escherichia coli* and *Fusobacterium nucleatum* have been shown to worsen inflammation in mouse models of tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @kostic_fusobacterium_2013]. In contrast to the patchiness of the taxa that were positively associated with carcinomas, potentially beneficial taxa had a more consistent association [Figure 6]. This result was particularly interesting because members of these taxa (i.e. *Ruminococcus* and *Clostridium XI* in stool and *Dorea* and *Blautia* in tissue) are thought to be beneficial due to their involvement in production of anti-inflammatory short chain fatty acids [@Venkataraman2016; @Herrmann2018; @Reichardt2017].

All of the adenoma classification models performed poorly. This result is not inconsistent with previous studies [@hale_shifts_2017; @Sze2017]. However, the classification results are at odds with results of the multitarget microbiota test (MMT) from Baxter, *et al.* [@baxter_microbiota-based_2016] who observed an AUC of 0.755 when applied to individuals with adenomas. There are two major differences between the models generated in this meta-analysis and that analysis. The MMT attempted to classify individuals as having a normal colon or having colonic lesions (i.e. adenomas or carcinomas) and not adenomas alone. Further, the MMT incorporated fecal immunoglobulin test (FIT) data while our models only used 16S rRNA gene sequencing data. Because FIT data were not available for the other studies in our meta-analysis, it was not possible to validate the MMT approach. The ability to differentiate between individuals with and without adenomas is an important problem since early detection of tumors is critical. However, it is possible that we might have been able to detect differences in the bacterial community if individuals with non-advanced and advanced adenomas were separated. This is a clinically relevant distinction since advanced adenomas are at highest risk of progressing to a carcinoma. The initial changes of the microbiota during tumorigenesis could be focal to where the initial adenoma develops and would not be easily assessed using fecal samples from an individual with non-advanced adenomas. Unfortunately, distinguishing between individuals with advanced and non-advanced adenomas was not possible in our meta-analysis since the studies did not provide the clinical data needed to make that distinction.

Stool samples represent a non-invasive approach to assessing the structure of the gut microbiota and are potentially useful for diagnosing individuals as having colonic tumors. However, they do not reflect the structure of the mucosal microbiota [@Flynn_preprint_2017]. Regardless, the taxa that were the most important in the stool-based models overlapped with those from the models trained using the data from unmatched and matched colon tissue samples [Figure S3]. Mucosal biopsies are preferred for focused mechanistic studies and have offered researchers the opportunity to sample healthy and diseased tissue from the same individuals (i.e. matched) using each individual as their own control or in a cross sectional design (i.e. unmatched). Because obtaining these samples is invasive, carries risks to the individual, and is expensive, studies investigating the structure of the mucosal microbiota generally have a limited number of participants. Thus, it was not surprising that tissue-based studies did not provide clearer associations between the mucosal microbiota and the presence of tumors. Interestingly, Fusobacterium, which has received increased recent attention for its potential role in tumorigenesis [@kostic_fusobacterium_2013] was not consistently identified across the studies in our meta-analysis. This could be due to the relatively small number of individuals in the limited number of studies. The classification models trained using the tissue-based data performed well when tested with the training data (Figure S4), but performed poorly when tested on the other tissue-associated datasets (Figure S5). Disturbingly, taxa that are commonly associated with reagent contamination (e.g. *Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc.) were detected within the tissue datasets. Such contamination is common in studies where there is relatively low bacterial biomass [@Salter_contamination_2014]. The lack of replication among the tissue-based biomarkers may be a product of the relatively small number of studies and individuals per study and possible reagent contamination.

Among our stool samples, we failed to identify several notable populations that are commonly associated with carcinomas including an enterotoxigenic strain of *Bacteroides fragilis* (ETBF) and *Streptococcus gallolyticus* subsp. *gallolyticus* [@bfrag_Boleij_2014; @strep_Aymeric_2017]. ETBF have been found in tumors in the proximal colon where they tend to form biofilms [@Purcell2017; @dejea_microbiota_2014]. Considering DNA from bacteria that are more prevalent in the proximal colon may be degraded by the time it leaves the body, it is not surprising that we failed to identify a significant OR for Bacteroides with carcinomas. In addition, since our approach could only classify sequences to the genus level and there are likely multiple Bacteroides populations in the colon, it is possible that sequences from ETBF and non-oncogenic Bacteroides were pooled. This would then reduce the OR between Bacterioides and whether an individual had carcinomas. It is also necessary to distinguish between populations that are biomarkers for a disease and those that are known to cause disease. The former may also have a causative role in the disease. Although the latter have been shown to have a causative role, they may appear at low relative abundance, be found in specific locations, or may have a highly patchy distribution among affected individuals.

Meta-analyses are a useful tool in microbiome research because they can demonstrate whether a result can be replicated and facilitate new discoveries by pooling multiple independent investigations. There have been several meta-analyses similar to this study that have sought biomarkers for obesity [@Sze2016; @Walters2014; @Finucane2014], inflammatory bowel disease [@Walters2014], and colorectal cancer [@shah_leveraging_2017]. Considering microbiome research is particularly prone to hype and overgeneralization of results, these analyses are critical. For example, previous meta-analyses have demonstrated that there are no clear fecal biomarkers for obesity [Sze2016; @Walters2014; @Finucane2014]. Such meta-analyses are difficult to perform because the underlying 16S rRNA gene sequence data are not publicly available, metadata are missing, incomplete, or vague, sequence data are of poor quality or derived by non-standard approaches, and the original studies were significantly underpowered. Reluctance to publish negative results (i.e. the “file drawer effect”) is also likely to skew our understanding of the relationship between microbiota and disease. Better attention to these specifics will increase the reproducibility and replicability of microbiota studies and make it easier to perform these crucial meta-analyses. Moving forward, meta-analyses will be important tools to help aggregate and find commonalities across studies when investigating the microbiota in the context of a specific disease [@Sze2016; @Walters2014; @Finucane2014; @shah_leveraging_2017].

Our meta-analysis suggests a strong association between the gut microbiota and colon tumorigenesis. By aggregating the results from studies that sequenced the 16S rRNA gene from fecal and tissue samples, we are able to provide evidence supporting the use of microbial biomarkers to diagnose the presence of colonic tumors. Further development of microbial biomarkers should focus on including other biomarkers (e.g. FIT), better categorizing of people with adenomas, and expanding datasets to include larger numbers of individuals. Based on prior research into the physiology of the biomarkers we identified, it is likely that they have a causative role in tumorigenesis. Their patchy distribution across individuals suggests that there are either multiple mechanisms causing disease or a single mechanism (e.g. inflammation) that can be mediated by multiple, diverse bacteria.

@@

***Random Forest classification analysis:*** 
To classify individuals as having normal colons or tumors, we built Random Forest classification models for each dataset and comparison using taxa with significant ORs (after multiple comparison correction), all taxa, or OTUs. Because no taxa were identified as having a significant OR associated with adenomas using stool samples or tissue samples, classification models based on OR data were not constructed to classify individuals as having normal colons or adenomas. Within the training dataset, 10-fold cross validation (5-fold cross validation for small datasets) was used to build a model that was then evaluated on the testing set. For the models constructed using the taxa with significant ORs, the default mtry setting was used to train the model and this model was tested on the other datasets in the meta-analysis. The reported AUC values are the AUCs for the application of the model on the test sets. For the OTU-based models, the dataset was split into training (80% of samples) and testing (20%) sets and 10-fold cross validation (5-fold cross validation for small datasets) on the training set was used to generate the model for the testing set. The original 80/20 split and fitting was repeated 100 times and the average AUC from these 100 repeats was reported. The Mean Decrease in Accuracy (MDA), a measure of the importance of each taxon to the overall model was used to rank the taxa used in each model. For all models, the default setting used was $\sqrt{p}$, where p is all the variables used in the respective model. Normally, $\sqrt{p}$ has been found to be what is chosen as the ideal mtry [@Breiman2001].
